---
title: "Dirichlet process Gaussian mixture model for clustering satellite images"
author: "Insha Ullah"
date: "March 15, 2018"
output:
  html_document: default
  pdf_document: default
---

Some required R packages. 
```{r}
library(MCMCpack)
# For parallel computation if you use Linux system you will need the following doMC package
#library(doMC) 
# For parallel computation if you use Windows you will need the following doSNOW package
library(doSNOW) 
# For parallel computation on both Linux and Windows you will need the following foreach package
library(foreach) 
library(plyr)
```



Following are the functions that will be called in the main function below to conduct MCMC.

This function samples alpha: the Dirichlet concentration parameter. 
```{r}
sample_alpha <- function(kmax,aalpha,balpha,V) {
  alpha=0
  while(alpha==0){
    alpha=rgamma(1,shape=aalpha+kmax-1,rate=balpha-sum(log(1-V[1:(kmax-1)])))
  }
  return(alpha)
}
```

```{r}
# This fuction is to sample the labels z and the mixing proportions pi=(pi_1,...,pi_k).
# The function take data x (number of variables * number of observations), 
# maximum number of clusters as kmax, clusters probabilities as pi0, 
# mean mu (number of variables * number of clusters), 
# covariance Sigma (number of variables * number of variables * number of clusters),
# and alpha: the Dirichlet concentration parameter.
# Note the variables are in rows of x and the observations are in columns.

sample_zpidp <- function(x,kmax,pi0,mu,Sigma,alpha){
  p <- nrow(x)
  n <- ncol(x)
  # first sample indicators for all data points:
  lpi=log(pi0)
  
  probx=array(NA,dim=c(kmax,n)); lpi=log(pi0);
  
  for (j in 1:kmax){
    probx[j,]= lpi[j]+logmvnormpdf(x, mu[, j],Sigma[ , ,j]);
  } 
  
  probx=exp(probx-matrix(apply(probx,2,max),kmax,n,byrow=TRUE))
  probx=probx/(matrix(apply(probx,2,sum),kmax,n,byrow=TRUE))
  
  z <- 1+apply(apply(probx,2,cumsum) <= matrix(runif(n),kmax,n,byrow=TRUE),2,sum); # Mn(1,probx)
  nz <- apply(t(outer(z,1:kmax,'==')),1,sum)
  
  # This is bit is to deal with the label switching issue
  snz <- sort(nz, decreasing = TRUE, index.return=TRUE)
  nz <- snz$x
  oldz=z;
  for (j in 1:kmax){
    z[which(snz$ix[j]==oldz)]=j;
  }
  
  # number of clusters
  k <- sum(nz!=0);
  
  
  # then sample the pi probabilities
  V=rep(0,kmax);
  npi=rep(0,kmax);
  
  for (j in 1:(kmax-1)){
    gamma1 <- 1+nz[j];
    gamma2 <- alpha + sum(nz[(j+1):kmax]);
    V[j]=1;
    while(V[j]==1){
      V[j]=rbeta(1,gamma1,gamma2);
    }
    
    if (j==1){
      npi[j] <- V[j]
    } else {npi[j] <- V[j]*prod(1-V[1:(j-1)]);}
    
  }
  V[kmax] <- 1;
  npi[kmax] <- prod(1-V[1:(kmax-1)])
  
  
  
  results <- list()
  results$z <- z
  results$V <- V
  results$pi0 <- npi
  results$k <- k
  return(results)
}
```

The following function is to sample mean mu and covariance Sigma.

```{r}
sample_muSigma <- function(x,p,kmax,z,m0,t0,s0,S0, mu, Sigma){
  #
  # sample posterior for parameters of k normals given config indicators
  # reference prior ...
  
  for (j in 1:kmax){
    
    i=which(z==j); 
    nj=length(i); 
    if (nj > 0) {
      a = t0/(1+nj*t0); 
      m = matrix(apply(x[,i,drop=FALSE],1,mean),p,1); 
      e= x[,i,drop=FALSE]-matrix(m,p,nj); 
      m=m-m0;
      S = e%*%t(e)+m%*%t(m)*nj*a/t0; 
      Sigma[ , , j] = riwish(s0+p-1+nj,S0+S);
      mu[ , j] = a*nj*m+sqrt(a)*t(chol(Sigma[ , , j]))%*%matrix(rnorm(p),p,1); 
    } else {
      Sigma[ , , j] = riwish(s0+p-1,S0);
      mu[ , j] = sqrt(t0)*t(chol(Sigma[ , , j]))%*%matrix(rnorm(p),p,1); 
    }
  }
  
  
  results <- list()
  results$mu <- mu
  results$Sigma <- Sigma
  return(results)
}
```

The following function returns log likelihood of the of the observation given the parameters mu and Sigma.

``` {r}
logmvnormpdf <- function(x,mu,Sigma){
  
  # x  = p.n array of n values of p-dim MV normal
  # mu = column p vector mean
  # Sigma = p.p variance matrix 
  # logpdf = n vector of log pdf values
  #
  p <- nrow(x)
  n <- ncol(x)
  C=chol(Sigma); e=t(solve(C))%*%(x-matrix(mu,p,n)); 
  if (p==1){
    q <- e^2
  } else {q <- apply(e^2,2,sum)}
  logpdf = -q/2 - sum(log(diag(C))) - log(2*base::pi)*(p/2);
  return(logpdf)
}

```

function to find mode.
```{r}

getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

```


```{r}
#update the stick-breaking parameters, v
SB.V <- function(kmax,gamma1,gamma2){
  V <- rep(1,kmax)
  for(j in 1:(kmax-1)){
    V[j] <- rbeta(1,gamma1,gamma2)
  }
  V
}
```

```{r}
#compute the stick-breaking weights
pi_weights <- function(V){ 
  kmax <- length(V)
  probs <- V
  probs[2:kmax] <- probs[2:kmax]*cumprod(1-V[2:kmax-1])
  probs
}
```

Blocked gibbs sampler.
```{r}
block_gibbs <- function(x){
  
  n <- dim(x)[2]
  p <- dim(x)[1]
  
  nburnin <- 1000             #burn-in iterations
  rsnmc <- 2000                 #mcmc iterations (post burn-in)
  
  k=30;
  kmax=30 #maximum number of components in truncated dirichlet
  
  #initialize:
  aalpha=0.0001; #dirichlet alpha priors
  balpha=0.0001;
  
  m0=matrix(0,p,1); t0=15; s0=5; S0=s0*diag(p)/2;
  
  
  V <- SB.V(kmax,gamma1=1,gamma2=1)
  pi0 <- pi_weights(V)
  
  
  mu <- array(NA,dim=c(p,k)); Sigma=array(NA,dim=c(p,p,k));
  for (j in 1:k){
    Sigma[ , , j]=S0/s0;
    mu[ , j] = 0.2*sqrt(t0)*t(chol(Sigma[ , , j]))%*%matrix(rnorm(p),p,1);
  }
  
  
  #save thinned iterations:
  mutotal <- array(0,dim=c(p,kmax,rsnmc));
  Sigmatotal=array(0,dim=c(p,p,kmax,rsnmc));
  pitotal=array(0,dim=c(kmax,rsnmc));
  Vtotal=array(0,dim=c(kmax,rsnmc));
  alphatotal=array(0,dim=c(1,rsnmc));
  ktotal=array(0,dim=c(1,rsnmc));
  ztotal=array(0,dim=c(ncol(x),rsnmc));
  zsorted=array(0,dim=c(1,ncol(x)));
  alpha=1;
  
  for (it in 1:(nburnin+rsnmc)){
    if(it%%500==0){
      cat("it=",it,"\n")
    }
    
    zpidp <- sample_zpidp(x,kmax,pi0,mu,Sigma,alpha); 
    z <- zpidp$z
    pi0 <- zpidp$pi0
    V <- zpidp$V
    k <- zpidp$k
    
    nz <- apply(outer(z,1:kmax,'=='),2,sum)
    
    alpha <- sample_alpha(kmax,aalpha,balpha,V);
    
    muSigma <- sample_muSigma(x,p,kmax,z,m0,t0,s0,S0,mu,Sigma);
    mu <- muSigma$mu
    Sigma <- muSigma$Sigma
    
    if(it > nburnin){
      saveit <- it-nburnin;
      mutotal[ , , saveit] <- mu;
      Sigmatotal[ , , ,saveit] <- Sigma;
      ktotal[ , saveit] <- k;
      alphatotal[ , saveit] <- alpha;
      pitotal[ , saveit] <- pi0;
      Vtotal[ , saveit] <- V;
      ztotal[ , saveit] <- z;
    }
  }
  
  results <- list()
  results$z <- apply(ztotal,1,getmode)
  results$Kpost <- ktotal
  results$alphapost <- alphatotal
  results$mupost <- mutotal
  results$Sigmapost <- Sigmatotal
  results$pipost <- pitotal
  results$Vpost <- Vtotal
  return(results)
}

```

Here we call the above functions 

DPGMM_res <- block_gibbs(t(yy))
```
